---
title: "Reproducible Research Project 2 - US NOAA Storm Database: health and economic
  consequences of severe weather"
author: "Brandon Robinson"
date: "`r Sys.Date()`"
output: html_document
editor_options: 
  markdown: 
    wrap: 80
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Synopsis

The second project of the Coursera course Reproducible Data, requires the
student to analyze the health and economic consequences of severe weather
between the years 1950 - 2011. The data to be analyzed will be provided by US
NOAA database which tracks weather events across the US. The questions to answer
are as follows:

1.  Across the United States, which types of events (as indicated by the
    *EVTYPE* variable) are most harmful with respect to population health?
2.  Across the United States, which types of events have the greatest economic
    consequences?

## Loading and Pre-Processing the Data

First the following packages will be loaded before running further code

```{r, echo = TRUE} 
library(R.utils)
library(data.table) 
library(dplyr) 
library(ggplot2) 
library(tidyr)
```
Next, the storm data file is downloaded and stored

```{r, echo = TRUE} 
if (!file.exists("stormdata.csv.bz2")) { 
        fileURL <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2"
        download.file(fileURL, "stormdata.csv.bz2")
        bunzip2("stormdata.csv.bz2", "stormdata.csv", remove=FALSE)
}
storm <- data.table::fread("stormdata.csv", fill=TRUE, header=TRUE) 
head(storm)
```

Review the column names and extract only the required names for analysis 
```{r, echo = TRUE}  
names(storm)
stormdata <- storm %>% select(c("EVTYPE","FATALITIES","INJURIES","PROPDMG","PROPDMGEXP","CROPDMG","CROPDMGEXP"))
str(stormdata)
```

The remaining variables for analysis are: 

1. EVTYPE - storm event type 
2. FATALITIES - fatalities per event 
3. INJURIES - injuries per event 
4. PROPDMG -property damage cost 
5. PROPDMGEXP - property damage exponent 
6. CROPDMG - cropdamage cost 
7. CROPDMGEXP - crop damage exponent

## Processing Data

The next step is to process the data based on the questions to be answered. The
first question asks which storm events are most harmful with respect to
population health.

### Question 1

The data will be processed by grouping event type with the total injuries and
deaths. Then, the data will be grouped in descending order to clearly delineate
the most harmful storm type. The goal is to get the data on the same plot for
easier comparison

```{r, echo = TRUE} 
totalInjuries <- aggregate(INJURIES ~ EVTYPE, data = stormdata, FUN = sum)
topInjuries <- totalInjuries[order(-totalInjuries$INJURIES), ] 
totalFatalities <- aggregate(FATALITIES ~ EVTYPE, data = stormdata, FUN = sum)
topFatalities <- totalFatalities[order(-totalFatalities$FATALITIES), ]

#### combine the injury and fatality data to one table and order the top 10

topHealth <- merge(topInjuries, topFatalities, all.x = TRUE)
topHealth[is.na(topHealth)] <- 0
topHealth2 <- topHealth[order(-topHealth$INJURIES),][1:10,]
topHealth2 <- gather(topHealth2,key=type, value=value, INJURIES,FATALITIES)
```

### Question 2

The second question asks which storm types have the greatest economic
consequences. The methodology to determine the economic impact will be similar
to the injuries and fatalities, but with a few extra steps to aggregate the
costs. The data will be combined in order to be viewable on the same plot.

The first step is to create a function to set the property damage exponents to
real numbers. This will allow us to collect the damage costs to property and
crops 

```{r, echo = TRUE}
cost <- function(x) {
        if (x == "H")
                100
        else if (x == "K")
                1000
        else if (x == "M")
                1e+06
        else if (x == "B")
                1e+09
        else if (x == "")
                1
        else if (x == "m")
                1e+06
        else if (x == "0")
                1
        else if (x == "5")
                1e+05
        else if (x == "6")
                1e+06
        else if (x == "4")
                10000
        else if (x == "2")
                100
        else if (x == "3")
                1000
        else if (x == "h")
                100
        else if (x == "7")
                1e+07
        else if (x == "1")
                10
        else if (x == "8")
                1e+08
        else
                0
}
```

Next the data will be processed by grouping event type with property and crop
damage costs. Then, the data will be grouped in descending order to clearly
delineate the greatest economic impact.

```{r, echo = TRUE} 
econdata <- storm %>% select(c("EVTYPE","PROPDMG","PROPDMGEXP","CROPDMG","CROPDMGEXP"))
econdata$Property <-  econdata$PROPDMG*sapply(econdata$PROPDMGEXP, FUN = cost)
econdata$Crops <- econdata$CROPDMG*sapply(econdata$CROPDMGEXP, FUN = cost)
dmgValue <- aggregate(Property~EVTYPE,data = econdata, FUN = sum, na.rm = TRUE)
dmgValue <- dmgValue[with(dmgValue,order(-Property)),]
cropValue <- aggregate(Crops~EVTYPE,data = econdata, FUN = sum, na.rm = TRUE)
cropValue <- cropValue[with(cropValue,order(-Crops)),]

#### combine economic losses into one table and order the top 10

topDmg <- merge(dmgValue, cropValue, all.x = TRUE)
topDmg[is.na(topDmg)] <- 0
topDmg2 <- topDmg[order(-topDmg$Property),][1:10,]
topDmg2 <- gather(topDmg2,key=type, value=value, Property,Crops)
```

## Results and Conclusions

The first question asked for the events which caused the most fatalities and
injuries. The below plot represents the top 10 storm events which were most
harmful to population health

```{r, echo = TRUE} 
damagePlot <- ggplot(data=topHealth2, aes(reorder(EVTYPE, -value), value, fill=type)) +
        geom_bar(position = "dodge", stat="identity") + 
        labs(x="Storm Type", y="Count") +
        theme(axis.text.x = element_text(angle = 45, vjust=0.5)) +
        ggtitle("Total Number of Injuries and Fatalities of the top 10 storm types")
print(damagePlot)
```


Based on the plot, Tornados had the greatest impact on population health.

The second question asked for the storm events which had the greatest economic
consequences. The below plot represents the top 10 storm events with the
greatest economic consequences.

```{r, echo = TRUE}  
costPlot <- ggplot(data=topDmg2, aes(reorder(EVTYPE, -value), value, fill=type)) +
        geom_bar(position = "dodge", stat="identity") + 
        labs(x="Storm Type", y="Cost") +
        theme(axis.text.x = element_text(angle = 45, vjust=0.5)) +
        ggtitle("Total Property and Crop Damage Cost of the top 10 storm types")
print(costPlot)
```


Based on the plot, Floods had the greatest economic loss impact in
the US
